### 2023-03-07 chapters/conclusion.tex
conclusion

\begin{itemize}
	\item 
	\textbf{Post-hoc Explainability.} Black-box classifiers are hard to understand and trust, especially in safety-critical domains. In this thesis, we have designed interpretable classifiers that are explainable by design. We can use these classifiers, such as {\imli} and {\crr}, to generate post-hoc explanations for black-box classifiers. In this context, logic-based explanations, such as abduction-based explanations~\cite{ignatiev2019abduction}, Anchor~\cite{ribeiro2018anchors}, and LORE~\cite{guidotti2018local} have demonstrated to be effective in generating post-hoc explanations of black-box classifiers. Therefore, we plan to test how well {\imli} can synthesize explanations with large samples and how well {\crr} can generate more expressive explanations.

	\item \textbf{Fairness Auditing.} Machine learning has demonstrated significant success as a predictive system in automated decision-making. However, quantifying and mitigating bias in these systems is still challenging, e.g. bias in generative machine learning and large language models~\cite{abid2021persistent,nadeem2020stereoset,vig2020investigating}. As a result, there has been significant research interest in auditing classifiers for bias, from standard supervised learning to deep neural networks, computer vision, large language models and so on. Thus, we need a fairness auditing framework~\cite{ruf2021towards,yan2022active} with formal guarantees. One task of this framework is to categorize different definitions of fairness metrics in the literature. This would help stakeholders pick the right definition of fairness for their application~\cite{ruf2021towards}. In addition, we aim to extend formal fairness verification beyond group fairness, such as individual fairness~\cite{john2020verifying}, causal fairness~\cite{pan2021explaining,zhang2018fairness}, and counterfactual fairness~\cite{wu2019counterfactual,chiappa2019path}. We also plan to cover more classes of machine learning classifiers that are hard or impossible to encode as CNF, such as deep neural networks. Additionally, we aim to extend fairness verification beyond tabular data to more complex data such as images and texts.

	\item \textbf{Fairness and Interpretability: Bridging the Gap.} We aim to design fair and interpretable algorithms for  machine learning. We are interested in how these two goals relate to each other. We have some research questions for the future: (i) are interpretable models fair? (ii) how can we improve the fairness of interpretable models? (iii) how can we use interpretability to enhance fairness? (iv) how can we optimize a classifier for both fairness and interpretability? Our work on {\fairXplainer} is a stepping stone towards bridging the gap between fairness and interpretability. We plan to explore more directions in the future.

	\item \textbf{Efficiency through Incremental Solving.} We aim to apply incremental solving, which we used for interpretable classification learning with {\imli} and {\crr}, to other optimization problems, not only in machine learning. This can help us scale up without losing much accuracy. For example, we aim to apply incremental solving in online learning of interpretable models, such as decision trees, decision lists, and decision sets. To conclude, the long-term vision of this thesis is to enhance machine learning in interpretability, fairness, and beyond, with the help of formal methods and automated reasoning.
\end{itemize}

>>>>

\begin{itemize}
	\item 
	\textbf{Post-hoc Explainability.} Black-box classifiers are hard to understand and trust, especially in safety-critical domains. In this thesis, we have designed interpretable classifiers that are explainable by design. A natural application of interpretable models is to apply them to explain the working of black-box classifiers. For example, rule-based classifiers can be leveraged to provide explanations in the form of classification rules, such as abduction-based explanations~\cite{ignatiev2019abduction}, Anchor~\cite{ribeiro2018anchors}, and LORE~\cite{guidotti2018local}. As such, we can use interpretable classifiers {\imli} and {\crr} to generate post-hoc explanations for black-box classifiers.  In future, we plan to test how well {\imli} can synthesize explanations with large samples and how well {\crr} can generate more expressive explanations.

	\item \textbf{Efficiency through Incremental Solving.} We aim to apply incremental solving, which we used for interpretable classification learning with {\imli} and {\crr}, to other optimization problems, even beyond machine learning. This can help us scale up solutions without losing much accuracy. For example, we aim to apply incremental solving in online learning of interpretable models, such as decision trees, decision lists, and decision sets, where the classifier can be routinely updated with new samples. Furthermore, we aim to try different objective functions in the incremental solving, apart from the hamming distance-based heuristic objective, which  works well in the incremental learning of CNF classification rules. One idea is to learn different rule-based classifiers, possibly in parallel, on different mini-batches or on different subsets of features, and then construct a rule merging procedure to combine different classification rules into one.

	\item \textbf{Fairness Auditing.} Any technology that is publicly used presently undergoes an audit mechanism, where we understand the impacts and limitations of using that technology, and why are they caused. Machine learning is becoming the pervasive technology of our time and the discourse on bias induced by machine learning systems is attracting attention, e.g. the demonstration of bias in popular generative machine learning and large language models~\cite{abid2021persistent,nadeem2020stereoset,vig2020investigating}. As a result, there has been significant research interest in auditing classifiers for bias, from standard supervised learning to deep neural networks, computer vision, large language models and so on. Thus, we aim to design a fairness auditing framework~\cite{ruf2021towards,yan2022active} with formal guarantees.  There are three key questions that we aim to investigate in fairness auditing.

	\begin{enumerate}
		\item \textbf{Which fairness metrics to choose?} Fairness in machine learning is bestowed with multiple notions of fairness. Our first line of investigation is to categorize different fairness metrics  and suggest the best metric based on application, data, and prevailing policy, similar to ``Fairness Compass''~\cite{ruf2021towards}. This would help stakeholders pick the right definition of fairness for their application.

		\item \textbf{How to quantify bias?} Accurate quantification of bias is an important step towards designing algorithms to mitigate bias. As discussed in the thesis, fairness verification allows us to formally quantify the bias of a classifier. To this end, we aim to extend formal fairness verification to broader classes of fairness metrics, classifiers, and data.

		\textbf{Fairness Metrics.} We aim to extend fairness verification of beyond group fairness, such as individual fairness~\cite{john2020verifying}, causal fairness~\cite{pan2021explaining,zhang2018fairness}, and counterfactual fairness~\cite{wu2019counterfactual,chiappa2019path}. Each category of fairness metrics poses distinct challenges in formal verification. For example, verifying individual fairness is related to verifying robustness of a model. As such, statistical methods from robustness verification has been applied to individual fairness~\cite{john2020verifying}. Formal methods, such as SMT-based encoding is also proposed in this regard~\cite{biswas2022fairify}. In our research endeavor, we aim to leverage  SAT or quantified Boolean formula (QBF) based verification, which may improve the scalability of verification.

		\textbf{Broader Classifiers.} We aim to extend fairness verification to broader machine learning classifiers, such as random forests and deep neural networks. For random forests, we can leverage CNF-based translation of the ensemble of trees by converting each tree as a CNF and a cardinality constraint to implement the ranking function of the prediction of multiple trees. For a special case of binarized neural networks (BNNs)~\cite{hubara2016binarized}, we can leverage existing CNF encoding and deploy SSAT-based verifier. However, for general neural networks with continuous parameters, MILP-based encoding can possibly be explored~\cite{mistry2022milp}. In addition, for neural networks, other surrogate representation beyond CNF or MILP can also be studied in future. In all cases, fairness verification, particularly for group-based metrics, relies on an access to efficient counter of CNF/MILP encoding; thus a dedicated effort to design better counter is a challenging, yet important research direction to explore.

		\textbf{Complex data.} We have explored fairness verification of tabular data; an important research question is to formally verify the fairness of classifiers with complex data such as images~\cite{nuriel2021permuted} and languages~\cite{abid2021persistent,nadeem2020stereoset,vig2020investigating}. For such data, one way to adapt existing verification methods, such as {\justicia}, is to apply them on the learned feature representation by the neural network and propagate verification results back to the input layer with images or text. In future, we aim to explore this possibility.

		\item \textbf{How to explain bias?} We aim to design fair and interpretable algorithms for  machine learning. We are interested in how these two goals relate to each other. This is inline with GDPR's emphasis on making models transparent and trustworthy, which is deployed in public.	Towards bridging the gap between fairness and interpretability in predictive systems, we explore following research questions:

		\begin{enumerate}
			\item How fair are interpretable machine learning models? 
			\item How can we improve the fairness of interpretable models?
			\item How can we apply interpretability to enhance fairness?
			\item How can we jointly optimize a classifier for  fairness and interpretability?
		\end{enumerate}

		Our work on {\fairXplainer} is a stepping stone towards bridging the gap between fairness and interpretability. We plan to explore more directions in the future.

	\end{enumerate}

\end{itemize}

To conclude, the long-term vision of this thesis is to enhance machine learning in interpretability, fairness, and beyond. To achieve this goal, we aim to integrate formal methods and automated reasoning with statistics and algorithm design.
